Analysis-ready VCF at Biobank-scale using Zarr

### 2024-05-08 paper.tex
Update caption for fig1

VCF is row-wise format in which 
describes observations
and metadata for a single variant~\citep{danecek2011variant} are
encoded as a line of text.
BCF (a packed binary representation) is similarly 
row-wise, as are the majority of proposed alternative storage formats.
Row-wise storage makes retrieving all information
for a given record straightforward and efficient, and 
allows data to be easily streamed through Unix pipelines (see Discussion).
The row-wise approach works well when records are relatively small, 
or when we typically want to analyse each record in its entirity.
When we want to analyse only a subset of a record,
row-wise storage can be inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for a set of rows means retrieving and decompressing \emph{all} 
information from these rows.

The usual alternative to row-wise storage is \emph{columnar} storage:
instead of grouping together all the fields for a record,
we group together all the records for a given field.
Columnar storage formats such as Parquet~\citep{parquet2024},
make retrieving particular columns much 
more efficient and can lead to substantially better compression.
While columnar techniques have been successfully applied 
in alignment
% TODO check
% Jeff: OK with the ADAM reference in here? You did some parquet based
% alignment storage, right?
storage~\citep[e.g.][]{bonfield2014scramble,nothaft2015rethinking,bonfield2022cram},
the use of technologies such as Parquet for 
storing and analysing variation data have had limited
success~\citep{boufea2017managing,fan2020variant}.
Columnar techologies tend to be 
optimised for ``tall and narrow'' data there is a large number of records
for a relatively small number of columns. 
Mapping VCF directly to a columnar layout, in which there is a 
a column for the genotypes (and other per-sample QC metrics) 
leads to millions of columns, and substantial technical challenges.
Additionally, compression performance within columns is poor 
because there is much less similarity among
the genotypes for a given sample than there is among the genotypes
for a given variant. [TODO check this - reference?]

\begin{figure}
\resizebox{225pt}{!}{\input{diagrams/data-model.tex}}
\caption{Columnar storage of VCF data in chunked, compressed form using Zarr.
[TODO more detail, including brief overview of how data maps from 
VCF to VCF Zarr in the examples]
\label{fig-data-model}}
\end{figure}

VCF is at its core an encoding of the genotype matrix where each entry
describes the observed genotypes for a given sample at a give variant site,
interleaved with per-variant information
and other matrices (e.g., the GQ or AD fields).
The data is largely numerical and each field is of a fixed dimension, 
and is therefore a natural mapping to array-oriented 
or ``tensor'' storage.
We propose the VCF Zarr specification which maps the 
VCF data model into an array-oriented layout using the
widely-used Zarr format (Fig~\ref{fig-data-model}). 
Briefly, Zarr is a format 
% FIXME two different uses of $n$ here in a paragraph
for storing $n$-dimensional arrays of a fixed datatype, stored as 
compressed chunks. In the VCF Zarr specification, 
each field in a VCF is mapped to an array, which is compressed 
and stored individually allowing for efficient retrieval. 
In particular, call-level data is stored as $m \times n$ arrays
(for $m$ sites and $n$ samples), allowing for efficient 
retrieval of subsets of those fields along both the 
variants and samples axis.
See the Methods for more detail on Zarr and the VCF Zarr
specification.

>>>>

VCF is row-wise format in which 
observations and metadata for a single variant are
encoded as a line of text~\citep{danecek2011variant}.
BCF (the standard binary representation of VCF) is similarly 
row-wise, as are the majority of proposed alternative storage formats.
Row-wise storage makes retrieving all information
for a given record straightforward and efficient. 
The row-wise approach works well when records are relatively small, 
or when we typically want to analyse each record in its entirity.
When we want to analyse only a subset of a record,
row-wise storage can be inefficient because we will usually need to
retrieve more information than required from storage. In the case 
of VCF (and BCF) where records are not of a fixed size and 
are almost always compressed in blocks, accessing any information
for a set of rows means retrieving and decompressing \emph{all} 
information from these rows.

The usual alternative to row-wise storage is \emph{columnar} storage:
instead of grouping together all the fields for a record,
we group together all the records for a given field.
Columnar storage formats such as Parquet~\citep{parquet2024}
make retrieving particular columns much 
more efficient and can lead to substantially better compression.
While columnar techniques have been successfully applied 
in alignment
% TODO check
% Jeff: OK with the ADAM reference in here? You did some parquet based
% alignment storage, right?
storage~\citep[e.g.][]{bonfield2014scramble,nothaft2015rethinking,bonfield2022cram},
the use of technologies such as Parquet for 
storing and analysing variation data have had limited
success~\citep{boufea2017managing,fan2020variant}.
Columnar techologies tend to be 
optimised for ``tall and narrow'' data where there is a large number of records
for a relatively small number of columns. 
Mapping VCF directly to a columnar layout, in which there is a 
column for the genotypes (and other per-call QC metrics) 
leads to millions of columns, and substantial technical challenges.
Additionally, compression performance within columns is poor 
because there is much less similarity among
the genotypes for a given sample than there is among the genotypes
for a given variant. [TODO check this - reference?]

\begin{figure}
\resizebox{225pt}{!}{\input{diagrams/data-model.tex}}
\caption{Chunked compressed storage of VCF data using Zarr. 
The \texttt{call\_genotype} array is a three-dimensional (variants, samples,
ploidy) array of integers, split into a uniform grid of 
chunks determined by the variant and sample chunk sizes (10,000
and 1,000 by default in \texttt{vcf2zarr}). Each chunk is associated 
with a key defining its location in this grid, which can be stored 
in any key-value store such as a standard file-system or cloud object
store. Chunks are compressed independently using standard 
codecs and pre-compression filters, which can be specified on a per-array
basis. Also shown are the one-dimensional \texttt{variant\_contig} (CHROM)
and \texttt{variant\_position} arrays (POS). Other fields are stored 
in a similar fashion. \label{fig-data-model}}
\end{figure}

VCF is at its core an encoding of the genotype matrix, where each entry
describes the observed genotypes for a given sample at a given variant site,
interleaved with per-variant information
and other call-level matrices (e.g., the GQ or AD fields).
The data is largely numerical and each field is of a fixed dimension, 
and is therefore a natural mapping to array-oriented 
or ``tensor'' storage.
We propose the VCF Zarr specification which maps the 
VCF data model into an array-oriented layout based on the
widely-used Zarr format (Fig~\ref{fig-data-model}). 
Briefly, Zarr stores multidimensional arrays of a given datatype
as compressed fixed-size chunks. In the VCF Zarr specification, 
each field in a VCF is mapped to an array, 
which is stored individually allowing for efficient retrieval and 
high levels of compression.
% In particular, call-level data is stored as $m \times n$ arrays
% (for $m$ sites and $n$ samples), allowing for efficient 
% retrieval of subsets of those fields along both the 
% variants and samples axis.
See the Methods for more detail on Zarr and the VCF Zarr
specification.
